---
title: "Optimizer Benchmark"
format:
  gfm:
    toc: false
knitr:
  opts_chunk:
    comment: "#>"
    message: false
    dev: "png"
    fig.width: 8
    fig.height: 4.5
    fig.align: center
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}

#' To install mytidyfunctions, you need 
#' remotes::install_github("JavierMtzRdz/mytidyfunctions")
if (!require("pacman")) install.packages("pacman")
if (!require("devtools")) install.packages("devtools")
if (require("mytidyfunctions")) mytidyfunctions::set_mytheme() 

Sys.setlocale("LC_ALL", "en_US.UTF-8")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
devtools::load_all(".")

pacman::p_load(survival, bench, ggplot2, dplyr, tidyr)

```

## Overview

This benchmark compares the three optimization algorithms available in `cbSCRIP`:

- **CCD**: Cyclical Coordinate Descent
- **SAGA**: Stochastic Average Gradient with Nesterov acceleration
- **SVRG**: Stochastic Variance Reduced Gradient

We evaluate performance across different dataset sizes measuring:

- **Execution time**
- **Memory allocation**

## Data Generation

```{r data-setup}
set.seed(42)

# Define benchmark configurations
configs <- tibble::tibble(
  n = c(200, 500, 1000),
  p = c(20, 50, 100),
  config_name = paste0("n=", n, ", p=", p)
)

# Generate datasets
datasets <- purrr::map2(configs$n, configs$p, function(n, p) {
  gen_data(n_train = n, p = p, num_true = 5, setting = 1)$train
})
names(datasets) <- configs$config_name
```

## Benchmark Function

```{r benchmark-fn}
run_optimizer_benchmark <- function(train_data, optimizer, nlambda = 10) {
  cbSCRIP(
    Surv(ftime, fstatus) ~ .,
    data = train_data,
    nlambda = nlambda,
    optimizer = optimizer,
    ratio = 20,
    maxit = 100,
    coeffs = "original"
  )
}
```

## Running Benchmarks

```{r run-benchmarks}
#| cache: true

optimizers <- c("CCD", "SAGA", "SVRG")

# Run benchmarks using bench::press to standardise output
benchmark_results <- bench::press(
  config_name = configs$config_name,
  {
    data <- datasets[[config_name]]
    bench::mark(
      CCD = run_optimizer_benchmark(data, "CCD"),
      SAGA = run_optimizer_benchmark(data, "SAGA"),
      SVRG = run_optimizer_benchmark(data, "SVRG"),
      iterations = 10,
      check = FALSE,
      memory = TRUE
    )
  }
)

# Ensure proper ordering of configurations
benchmark_results <- benchmark_results |>
  mutate(config_name = factor(config_name, levels = configs$config_name))

benchmark_results
```

## Results

### Execution Time

```{r time-plot}
#| fig-width: 10
#| fig-height: 6

autoplot(benchmark_results, type = "boxplot") +
  facet_wrap(~config_name, scales = "free_x") +
  labs(
    title = "Execution Time Distribution",
    subtitle = "Higher density / lower values are better",
    x = "Time"
  ) +
  theme(legend.position = "none")
```

### Memory Allocation

```{r memory-plot}
#| fig-width: 8
#| fig-height: 5

benchmark_results |>
  mutate(
    mem_mb = as.numeric(mem_alloc) / 1e6
  ) |>
  ggplot(aes(x = config_name, y = mem_mb, fill = as.character(expression))) +
  geom_col(position = position_dodge(width = 1)) +
  labs(
    title = "Memory Allocation by Optimizer",
    x = "Setting",
    y = "Memory Allocated (MB)",
    fill = "Optimizer"
  )
```

### Summary Table

```{r summary-table}

benchmark_results |>
  mutate(
    time_sec = round(as.numeric(median), 3),
    mem_mb = round(as.numeric(mem_alloc) / 1e6, 2)
  ) |>
  select(Setting = config_name, Optimizer = expression, 
         `Time (s)` = time_sec, `Memory (MB)` = mem_mb) |>
  knitr::kable()
```

## Session Info

```{r session-info}
sessionInfo()
```
